---
title: "Task 2.1 Unsupervised preprocessing and visualization"
author: "Johannes Schwietering"
date: "2024-04-05"
output: html_document
---
Load dataset food.csv


```{r setup}
knitr::opts_chunk$set(echo = TRUE)

root.dir = "/Users/johannesschwietering/Library/Mobile Documents/com~apple~CloudDocs/Studium/UMIT/Modul 12/AMLHC_Tasks_JS1"

# Load dataset food.csv
food_data <- read.csv("food.csv", header = TRUE, sep = ",")

```

## Get an overview by analyzing dimension, number of missing values and feature scaling


```{r overview}
# Analyze the dimensions of the dataset
dim_food_data <- dim(food_data) # Gets the dimensions of the data frame
cat("Dimensions of dataset: ", dim_food_data, "\n") # Prints out the dimensions

# Calculate the number of missing values in the dataset
num_missing_values <- sum(is.na(food_data)) # Counts the total number of NA values
cat("Number of missing values: ", num_missing_values, "\n") # Prints out the number of missing values

```

The results show that we have a dataframe with 25 rows and 10 columns. We ha no missing values. Die Z-Transformation hat funktioniert: Der Mittelwert ist für alle Featuers 0. Jetzt können wir die einzelnen Features sinnvoll miteinander vergleichen. 


## Preprocess data using z-transformation (R function scale)

```{r pressure}
# Feature scaling using z-transformation (standardization)
scaled_food_data <- scale(food_data[,-1])

# To check the scaling, let's look at the summary of the first few features
# This gives the mean and standard deviation which should be 0 and 1, respectively, after scaling
summary_scaled_food_data <- summary(scaled_food_data) 
print(summary_scaled_food_data) # Prints a summary statistics of the scaled data
```

## Perform a principal component analysis (R function princomp)

```{r principal component analysis}
# Perform PCA using the princomp function
pca_result <- princomp(scaled_food_data)

# Print a summary of the PCA results to get eigenvalues and the proportion of variance
summary(pca_result)

# View the loadings for the PCA components
loadings(pca_result)

# Eigenvalues can give us the variance explained by each principal component
pca_result$sdev^2

# To see the proportion of variance explained by the principal components
prop_var_explained <- pca_result$sdev^2 / sum(pca_result$sdev^2)
prop_var_explained
```


Create a score plot using PC1 and PC2 (R function qplot)
Update your git-repository. 

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
