---
title: "Task_3_1"
author: "Johannes Schwietering"
date: "2024-04-20"
output: html_document
---

# Unsupervised learning

##Loading Data

1) Load dataset food.csv

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)
# Load the necessary libraries for handling data
library(clusterCrit)
library(gplots)
library(fpc)


# Load the dataset named 'food.csv' with the first column as row names
food_data <- read.csv("food.csv", row.names=1)
# Display the first few rows of the dataset to verify it loaded correctly
head(food_data)

```

## Preprocessing
2) Preprocess data using z-transformation (R function scale)

```{r preprocess, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)

# Preprocess the data using z-transformation
# The scale function centers and scales the data (mean = 0, SD = 1)
# Applying z-transformation to all numerical columns in the dataset
scaled_food_data <- scale(food_data)

# Create a new data frame with the scaled data to handle further analyses if needed
scaled_food_data <- as.data.frame(scaled_food_data)

```

## Clustering
3) Apply KMeans for different number of clusters by varying k from 2 to 5. Calculate the Silhouette coefficient (function intCriteria of package clusterCrit) and select the final number of clusters

```{r KMeans, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)

# Convert the scaled data to matrix format required by the kmeans function
scaled_food_data <- as.matrix(scaled_food_data)

# Prepare to store Silhouette scores for comparison
silhouette_scores <- numeric(4)
k_values <- 2:5

# Set seed for reproducibility
set.seed(123)

# Iterate over k values from 2 to 5 to find the best k based on Silhouette score
for (k in k_values) {
  # Apply KMeans clustering
  kmc <- kmeans(scaled_food_data, centers = k)
  
  # Calculate the Silhouette coefficient
  cl <- intCriteria(scaled_food_data, kmc$cluster, "Silhouette")
  
  # Store the Silhouette score
  silhouette_scores[k - 1] <- cl
  
  # Print the cluster centers to inspect them
  print(paste("K =", k))
  print(head(kmc$centers))
}

# Print all Silhouette scores for each k
names(silhouette_scores) <- paste("k =", k_values)
print(silhouette_scores)

# Determine the best k based on the maximum Silhouette score
optimal_k <- which.max(silhouette_scores) + 1  # Adjust index by 1
cat("The optimal number of clusters (k) based on the Silhouette coefficient is:", optimal_k, "\n")

  kmc <- kmeans(scaled_food_data, centers = 2)

```

## PCA
4) Visualize the final kMeans clustering using PCA (R function princomp and qplot)

```{r Visualize, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)

# Perform PCA on the scaled data
pca_result <- princomp(scaled_food_data)

# Get PCA scores
scores <- pca_result$scores[,1:2]
scores <- scores*-1

# Plot using qplot from ggplot2
plot(scores[, "Comp.1"], scores[,"Comp.2"], col=kmc$cluster)
text(scores[,"Comp.1"], scores[,"Comp.2"], labels=rownames(scaled_food_data), cex=0.5)
```

## Hierarchical Clustering

5) Create a hierarchical clustering (R function hclust) and plot the dendrogram. In addition, create a heatmap with a clustering of samples and variables (R function heatmap)
```{r hierarchical, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)

# Compute hierarchical clustering on the distance matrix
  hcf <- hclust(dist(scaled_food_data))
  plot(hcf)
  
# Create a heatmap with dendrograms for both samples and variables
  heatmap.2(scaled_food_data, scale="none")
```

## Density based clustering
6) Create a density based clustering (R function dbscan of package fpc)
```{r clustering, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)

# Apply DBSCAN, a density-based clustering algorithm
  dbc <- dbscan(scaled_food_data,eps=2, MinPts=3)
  print(dbc)

# Plot the results with DBSCAN cluster assignments
  plot(scores[, "Comp.1"], scores[,"Comp.2"], col=dbc$cluster)
  text(scores[,"Comp.1"], scores[,"Comp.2"], labels=rownames(scaled_food_data), cex=0.5)

```


